---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SuCoNets

<!-- badges: start -->
<!-- badges: end -->

Nowadays, the way to create a co-expression network is to use Hierarchical Clustering. 
This package allows you to create a co-expression network using the glmnet algorithm. 
In addition, the network created will be relative to a covariate of the sample to be studied,
so we are creating a supervised coexpression network.

## Installation

You can install SuCoNets like so:

``` r
devtools::install_github("carmen-maria-hernandez/SuCoNets”)
```

## Example

Suppose we have an expression matrix, data, where the columns are blood samples and the rows are genes, so that each sample is identified by the numerical values taken by the genes. Let us also assume that the covariate we are going to study is the age of each individual to whom each blood sample corresponds. This covariate is given as a numerical vector, which we denote as age.

An example of a typical execution of the functions contained in this package would be as follows.


```{r, include=FALSE}
expr.data <- read.delim("~/WholeBlood.txt.gz",header = T,row.names = 1)
data.age.num <- read.csv("~/GTEx_Analysis_2017-06-05_v8_Annotations_SubjectPhenotypesDS.csv", header = T, sep = "\t")
data.age.num <- data.age.num[-980,]

rw <- data.age.num$SUBJID
rw <- gsub("-", ".", rw)

sample.prefixes10 <- substr(colnames(expr.data), 1,10)
sample.prefixes9 <- substr(colnames(expr.data), 1,9)

common.elements.length10 <- intersect(sample.prefixes10,rw)
common.elements.length9 <- intersect(sample.prefixes9,rw)

library(SuCoNets)

indexes.age.length10 <- selectRows(rw, common.elements.length10)
indexes.age.length9 <- selectRows(rw, common.elements.length9)

 indexes.samples.length10 <- selectRows(sample.prefixes10, common.elements.length10)
 indexes.samples.length9 <- selectRows(sample.prefixes9, common.elements.length9)

data.characteristic <- data.age.num[c(indexes.age.length10,indexes.age.length9),c(3,4,11)]

age = data.characteristic$AGE
data = expr.data[, c( indexes.samples.length10, indexes.samples.length9)]

nameGenes <- read.csv("~/protcodinggenes.txt", sep = " ")
nameGenesFileExpr <- rownames(data)
nameGenesFileExpr <- unlist(lapply(strsplit(nameGenesFileExpr, "\\."),function(x){ return(x[1])}))
genesBothFiles <- intersect(nameGenesFileExpr, nameGenes[,2])

ind.new.names <- selectRows(nameGenes[,2], genesBothFiles)
ind.old.names <-  selectRows(nameGenesFileExpr, genesBothFiles)

new.names <- nameGenes[ind.new.names,1]
data <- data[ind.old.names, ]
rownames(data) <- new.names
```

```{r, include = F}
rm(data.age.num,expr.data,common.elements.length10, common.elements.length9, rw, indexes.age.length10, indexes.age.length9,  indexes.samples.length10,  indexes.samples.length9, sample.prefixes10, sample.prefixes9, nameGenes, nameGenesFileExpr, genesBothFiles, ind.new.names, ind.old.names, new.names)
```


We start by loading the package *SuCoNets*. Then, we normalize the age with the function `normalize` and change to logarithmic scale, centralize and normalize the expression matrix data with the function `scn`. We also removed redundant predictors by invoking the function `rRedundantPredictors`.

Note that the function `normalize` returns: first the mean of the vector we pass as parameter, then its standard deviation and then the vector, which we pass as parameter, normalized.


```{r}
library(SuCoNets)
age <- normalize(age)
m <- age[1]
d <- age[2]
age <- age[-c(1,2)]
data <- scn(data)
data <- rRedundantPredictors(data)
```


Next, we calculate which seed produces the data partition (training set and test set) that gives the best results when running `glmnet` algorithm, so we can use that partition to run `glmnet` algorithm. With the function detectGenes we get the genes that `glmnet` algorithm has selected as important for age prediction


```{r}
seed <- bestSeed(data,age)

cvfit <- glmnetGenes(data,age, seed)
glmnet::print.cv.glmnet(cvfit)

selected.genes <- detectGenes(data,age,cvfit)

```

We can study the stability of the genes selected by the glmnet algorithm with the function `stabilitySelection`. This function runs the glmnet algorithm ten times by varying the input data set and saves the genes that in each run glmnet selects as important for predicting the covariate under study. As a result, for the genes we are going to work with, this function shows the number of times that each of them has been selected in the different executions. The interesting thing about this function is to see the statistics of this selection, calling the summary function to see the result of the execution. In addition, with the function `histGeneFreq` we can plot a histogram with the number of genes that appear once in the different runs, the genes that appear twice, etc. 


```{r}
selection.statistics <- stabilitySelection(data, age, selected.genes)
summary(selection.statistics[,2])
histGeneFreq(selection.statistics)
```


Now let's look at a scatter plot with the predicted age and the actual age of both the test set and the train set. Note that the mean and standard deviation of the age must be passed as parameters to this function to be able to see in the scatter plot the ages in the real ranges.


```{r}
comparisonActualPredictedCovariate(data, age, m, d, cvfit, seed)
```


We can also obtain a representation of all individuals in the analysis according to their age. Each individual will be represented by a dot. The blue dots represent the youngest individuals and the red dots represent the oldest. The individuals whose age is between these two, will be represented by a color contained in the gradient of these two colors. In order to perform this representation, we use the genes that the `glmnet` algorithm has selected, we perform a PCA on them and we plot the first two components of this PCA, which are the ones that explain the most variability in the data.


```{r}
distributionIndividualsCovariate (data, age, selected.genes, m,d)
```


Finally, we compute the supervised coexpression network. This package provides two methods for constructing the networks. The `coexpressionNetworkFixed` function builds each cluster of the network using a fixed size that we pass as a parameter. The `coexpressionNetworkVariable` function builds the clusters by adding genes incrementally until adding a new gene does not provide significant information. Thus, the size of each cluster is not necessarily the same. In addition, we can call function `running.gprofiler` to obtain biological information about the networks.


```{r}
network.fixed <- coexpressionNetworkFixed(data, selected.genes, 50)
network.variable <- coexpressionNetworkVariable(data, selected.genes, age, seed)

output.gprofiler2.fixed <- running.gprofiler(selected.genes, rep(51, nrow(selected.genes)), data, network.fixed)

output.gprofiler2.variable <- running.gprofiler(selected.genes, network.variable[[2]], data, network.variable[[1]])

```
