---
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "../../man/figures/Hub_genes_detection-",
  out.width = "100%"
)
```

# Hub Gene detection algorithm

In this tutorial, we will show the pipeline to extract the hub genes (or predictors) for a given covariate. These genes will be used to generate the co-expression networks.

## Requisites

The libraries required for this tutorial are `CovCoExpNets`, `magrittr` for the `%>%` pipe-like operator and `dplyr`. Additionally, `CovCoExpNets` uses `foreach` and `doParallel` to execute the functions in parallel.

```{r}
library(CovCoExpNets)
library(magrittr)
library(dplyr)
doParallel::registerDoParallel(13)
```

As for the input data, we will use the preprocessed data obtained from the [Data extraction tutorial](../Data_preparation). We need a data matrix with genes as rows and samples as columns, as well as a numeric vector for the covariate. In our case, we will use the data and age for the Cortex tissue, but this tutorial will work if a list of data matrices and a list of covariate vectors were used.

```{r}
brain_data.path <- "~/GTEx_data/"

data = readRDS(paste0(brain_data.path, "data.combined.rds"))
age = readRDS(paste0(brain_data.path, "age.combined.rds"))

m <- lapply(age, function(x) x$mean)
d <- lapply(age, function(x) x$standard.deviation)
age <- lapply(age, function(x) x$covariate)

data = data[["Cortex"]]
age = age[["Cortex"]]
m = m[["Cortex"]]
d = d[["Cortex"]]
```

## Summary

The whole pipeline to obtain the hub genes is as follows:

```{r cache = TRUE}
genes.freq = CovCoExpNets::geneFrequency(data, age, t = 10, seed = 1796)
genes.subset = CovCoExpNets::reduceGenes(genes.freq, mrfa = 0.9)
cvfit = CovCoExpNets::glmnetGenesSubset(data, age, genes.subset, seed = 1796)
genes.relevant = CovCoExpNets::extractModelGenes(cvfit)
```

The minimum requirements to execute the pipeline are the dataset and the covariate vector, the *t* hyperparameter (by default 10) and the *mrfa* hyperparameter (by defualt 0.9). The output is a list of the hub genes:

```{r}
# First 25 relevant genes for Cortex tissue:
genes.relevant$Genes[1:25]
```


## Steps

### Step 1: Executing the `glmnet` repetitions

The first step is to execute the `glmnet` repetitions a set number of times (`t`). To do so, we will use the `geneFrequency` function from `CovCoExpNets`.

```{r}
genes.freq = CovCoExpNets::geneFrequency(data, age, t = 10, seed = 1796)
```

As seen in the example, only the `data` and `covariate` variables are required by default. The result will be a data.frame with all genes selected by each iteration and their coefficients. 

```{r}
head(genes.freq, 5)
tail(genes.freq, 5)
```

### Step 2: Minimum relative ratio of appearance treshold

In this step, we reduce the list of predictors returned in each iteration. To do so, we use the hyperparameter *mrfa* to specify the minimum relative frequency in which a predictor must appear. To do so, we use the function `reduceGenes` from `CovCoExpNets`, where we count how many times each gene is selected as relevant across the `glmnet` repetitions and divide it by the total number of repetitions so that we have the percentage of repetitions in which it was selected as relevant. We require a minimum appearance threhsold. 

By default, we recommend an *mrfa* of 0.9: 

```{r}
genes.subset = CovCoExpNets::reduceGenes(genes.freq, mrfa = 0.9)

# First 25 genes:
genes.subset[1:25]
```

The output will be a set of genes that pass the threshold. In total, there were `r length(unique(genes.freq$Genes))` different genes across all repetitions, while `r length(genes.subset)` of those passed the threshold.

### Step 3: Model generation

Next, we create the final `glmnet` model. We use the `glmnetGenesSubset` function from `CovCoExpNets`.

```{r}
cvfit = CovCoExpNets::glmnetGenesSubset(data, age, genes.subset, seed = 1796)

cvfit
```

### Step 4: Genes and coefficients extraction

The final step is to extract the hub genes from the generated model. We use the `extractModelGenes` function from `glmnet`.

```{r}
genes.relevant = CovCoExpNets::extractModelGenes(cvfit)

head(genes.relevant, 10)
```

The list of hub genes is found in the "Genes" column of `genes.relevant` variables. The coefficients are also reported.

## Functions employed:

The required functions to execute the pipeline are:

* **geneFrequency:** executes the `glmnet` repetitions. It has the following arguments:

    * **data**: numeric data matrix with genes as rows and samples as columns. It also accepts a list of data matrices.
    * **covariate:** numeric vector with the age of the sample's donors. It also accepts a list of vectors.
    * **t:** number of `glmnet` repetitions to execute. Its value depends on the stability of the `glmnet` repetitions. Deaults to 10.
    * **k.folds:** (optional) number of k-folds in the `glmnet` cross-validation function `cv.glmnet`. Defaults to 10.
    * **train.split:** (optional) percentage to split the training dataset for testing purposes. A value of 0.8 will set the 80% of the samples to train and 20% to estimate the test RMSE. Defaults to 1.
    * **iter.RMSE:** (optional) whether to measure the RMSE of each reptition. Different solutions will not only have different returned predictors, but also different performance. Use this setting if you wish the prioritize better RMSE iterations when selecting the pruned predictors.
    * **data.test.extra:** (optional) numeric data matrix with genes as rows and samples as columns. If the train/test split was executed before this step and you want to estimate the RMSE over the test dataset, provide the test dataset in here. By default, none.
    * **covariate.test.extra:** (optional) numeric vector with the age of the sample's donors. If the train/test split was executed before this step and you want to estimate the RMSE over the test dataset, provide the test dataset in here. By defualt, none.
    * **sample.prob:** (optional) when executing the train/test split, it dictates the probability of each sample being executed. See [sample](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/sample) for more information.
    * **seed:** (optional) numeric seed to ensure reproducibility. Because of parallelization, even with an input seed results migh differ.
    * **glmnet.family:** (optional) family parameter used in the `cv.glmnet` function. See [`glmnet`](https://www.rdocumentation.org/packages/glmnet/versions/4.1-4/topics/cv.glmnet) for more details about the `family` parameter.

* **reduceGenes:** executes the minimum relative frequency of appearance (*mrfa*) threshold. It has the following arguments:

    * **genes.freq:** data.frame containing the genes that were selected in the `glmnet` repetitions. It must be obtained with the `geneFrequency` function from `CovCoExpNets`.
    * **mrfa:** numeric minimum relative frequency of appearance. If between
    * **force.mrfa:** (optional) whether to use only the *mrfa* parameter given. If set to FALSE, the parameter can be lowered until at least two predictors are selected. By default, TRUE.
    * **relative:** (optional) whether to use a relative *mrfa* or not. If TRUE, the parameter will be considered as a percentage from 0 to 1. If FALSE, the parameter will be considered the plain number of times the predictor must appear from 1 to t. By default, TRUE.

* **glmnetGenesSubset:** executes the final model training. It has the following arguments:
  
    * **data**: numeric data matrix with genes as rows and samples as columns. It also accepts a list of data matrices.
    * **covariate:** numeric vector with the age of the sample's donors. It also accepts a list of vectors.
    * **genes.subset:** character vector with the genes to be used to train the model. It must be the output of `reduceGenes` from `CovCoExpNets`. It also accepts a list of vectors.
    * **k.folds:** (optional) number of k-folds in the `glmnet` cross-validation function `cv.glmnet`. Defaults to 10.
    * **train.split:** (optional) percentage to split the training dataset for testing purposes. A value of 0.8 will set the 80% of the samples to train and 20% to estimate the test RMSE. Defaults to 1.
    * **sample.prob:** (optional) when executing the train/test split, it dictates the probability of each sample being executed. See [sample](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/sample) for more information.
    * **seed:** (optional) numeric seed to ensure reproducibility. Because of parallelization, even with an input seed results migh differ.
    * **glmnet.family:** (optional) family parameter used in the `cv.glmnet` function. See [`glmnet`](https://www.rdocumentation.org/packages/glmnet/versions/4.1-4/topics/cv.glmnet) for more details about the `family` parameter.
    * **return.genes.subset:** (optional) whether to return the hub genes along with the `glmnet` model. Recommended to FALSE and use `extractModelGenes` from `CovCoExpNets` over the model later on. Defaults to FALSE.

* **extractModelGenes:** extract the relevant predictors from a `glmnet` model. It has the following arguments.

    * **cvfit:** `glmnet` model to extract the relevant predictors and their coefficients from.
    * **genes.freq:** data.frame obtained from `geneFrequency`. If provided, it will also return the average coefficient of each gene across all glmnet repetitions. Defaults to none.

## Multiple conditions

As mentioned in the first section, we can input several covariates at the same time in the form of a list. The following pipeline will generate the hub genes for both Cerebellar Hemisphere and Cerebellum tissues.

```{r cache = TRUE}
data = readRDS(paste0(brain_data.path, "data.combined.rds"))
age = readRDS(paste0(brain_data.path, "age.combined.rds"))

m <- lapply(age, function(x) x$mean)
d <- lapply(age, function(x) x$standard.deviation)
age <- lapply(age, function(x) x$covariate)

data = data[c("Cerebellar Hemisphere", "Cerebellum")]
age = age[c("Cerebellar Hemisphere", "Cerebellum")]
m = m[c("Cerebellar Hemisphere", "Cerebellum")]
d = d[c("Cerebellar Hemisphere", "Cerebellum")]

genes.freq = CovCoExpNets::geneFrequency(data, age, t = 10, seed = 1796)
genes.subset = CovCoExpNets::reduceGenes(genes.freq, mrfa = 0.9)
cvfit = CovCoExpNets::glmnetGenesSubset(data, age, genes.subset, seed = 1796)
genes.relevant = CovCoExpNets::extractModelGenes(cvfit)
```

```{r cache = TRUE}
lapply(genes.relevant, function(x) head(x, 5))
```

