---
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "../../man/figures/Glmnet_stability_studies-",
  out.width = "100%"
)
```

# `Glmnet` stability studies

## Prerequisites

This tutorial requires for the libraries `magrittr` for its pipe-like operator, `dplyr`, `logger`, `jaccard` and `ggplot2`. We also automatically load the `doParallel` and `foreach` libraries.

```{r}
library(CovCoExpNets)
library(magrittr)
library(dplyr)
library(jaccard)
library(ggplot2)
library(logger)

#doParallel::registerDoParallel(13)
cl <- makeCluster(13)
doParallel::registerDoParallel(cl)
```

The file requirements are the data and age generated in the data extraction pipeline.

```{r}
brain_data.path <- "~/GTEx_data/"

data = readRDS(paste0(brain_data.path, "data.combined.rds"))
age = readRDS(paste0(brain_data.path, "age.combined.rds"))

m <- lapply(age, function(x) x$mean)
d <- lapply(age, function(x) x$standard.deviation)
age <- lapply(age, function(x) x$covariate)
```

## Steps

### Step 1: `Glmnet` repetitions

First, we will calculate the glmnet repetitions a total of *t=40* times, which should be enough to estimate with a high precision the stability.

```{r cache = TRUE}
genes.freq.40 = CovCoExpNets::geneFrequency(data, age, t = 40, seed = 1796)
```

### Step 2: Generation of the jaccard index matrix

Then, we generate a matrix with the repetitions as rows and columns, and the values of the intersection as the jaccard index between the two repetitions. We use the `measureStabilityJaccard` function:

```{r cache = TRUE}
glmnet.stability.matrix = CovCoExpNets::measureStabilityJaccard(genes.freq.40)
```

### Step 3: Stability results

We can calculate the average stability of the different tissues using the following function:

```{r cache = TRUE}
calculateAverageStability <- function(glmnet.stability.matrix, round.digits = 3){
  return.list = is(glmnet.stability.matrix, "list")
  if(!return.list){
    glmnet.stability.matrix = list(glmnet.stability.matrix)
  }
  
  df.stability = foreach(i = 1:length(glmnet.stability.matrix), .combine = "rbind") %do%{
    tissue.name = if(return.list) names(glmnet.stability.matrix)[i] else "Condition"
    
    df = CovCoExpNets::summaryVars(reshape2::melt(glmnet.stability.matrix[[i]], value.name = "stability"), variable = "stability", na.rm = T) %>%
      mutate(Tissue = tissue.name, .before = counter) %>% 
      select(Tissue, stability, ci) %>% mutate(across(where(is.numeric), round, round.digits))
    df
  }
  
  return(df.stability)
}

df.stability = calculateAverageStability(glmnet.stability.matrix, round.digits = 3)

df.stability
```

These results show the average jaccard index for every tissue. We can plot these results in barplots and compare the different tissues:

```{r fig.width=16, fig.height=9, dpi = 300, cache = TRUE}
plot = ggplot(df.stability, aes(x = reorder(Tissue, stability), y = stability)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", show.legend = T, fill = "steelblue") +
  labs(y = "Average jaccard index") + 
  theme(plot.title = element_text(size = 22, face = "bold"),
          panel.border = ggplot2::element_rect(colour = "black", fill = NA, size = 1),
          axis.text.x = ggplot2::element_text(face = "bold", color = "black", angle = 30, size = 15, hjust = 1),
          axis.text.y = ggplot2::element_text(face = "bold", color = "black", size = 15),
          axis.title.x = ggplot2::element_blank(),
          axis.title.y = ggplot2::element_text(face = "bold", size = 20, margin=margin(0,15,0,0)),
          panel.grid.minor = element_line(color = "gray", size = 0.1, linetype = 2),
          panel.grid.major = element_line(color = "gray", size = 0.2, linetype = 2),
          panel.background = element_rect(fill = "#F5F5F5"),
          aspect.ratio = 9/20,
          legend.title = element_text(size=20),
          legend.text = element_text(size=15),
          plot.margin = unit(c(1,1,1,1.5), "cm")
        )
plot
#ggsave("Stability_plots/Comparison.png", plot, width = 16, height = 9, dpi = 300, units = "in")
```


## Coincidence matrix:

We can also represent the heatplots with the coincidence matrices. This is an example for the Spinal cord tissue:

```{r fig.height=9, fig.width=10, dpi = 300}
genes.freq.10 = CovCoExpNets::geneFrequency(data$`Spinal cord (cervical c-1)`, age$`Spinal cord (cervical c-1)`, t = 10, seed = 1796)
plotAllSimilarity(genes.freq.10, diag = "genes", show.text = T)
```

We can also use the `output.path` argument to save the figure into the given location.

```{r}
stopCluster(cl)
```

