---
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "../../man/figures/Hub_genes_detection-",
  out.width = "100%"
)
```

# Covariate simulations

In this tutorial, we will study how to employ the simulation framework adapted from [simglm](https://cran.r-project.org/web/packages/simglm/index.html) and how to execute the `CovCoExpNets` to test the results.

## Requisites

The main library requirements for this tutorial are `CovCoExpNets`, `magrittr` for the pipe-like operator `%>%`, `dplyr`, `simglm` to generate the simulations and `caret` to evaluate the models. By default, the `CovCoExpNets` package also loads `foreach` and `doParallel` to execute the steps in parallel.

```{r}
library(CovCoExpNets)
library(magrittr)
library(dplyr)
library(simglm)
library(caret)

doParallel::registerDoParallel(13)
```

## Steps

### Step 1: Simulation generation

To execute a univariate simulation, we employ the `simulateUnivariate()` function from `CovCoExpNets`. The main parameters we need are the number of samples `N`, the total number of predictors `P` and how many of them will be relevant to the response variable `Q` (coefficients different from zero).

```{r}
N = 200
P = 1000
Q = 200

sim.obj = CovCoExpNets::simulateUnivariate(N, P, Q, seed = 1796)
```

The simulation object created is a list containing:

* **data:** the data.frame with the simulated dataset.
* **y:** the response variable.
* **df.coefficients:** a data.frame with two rows: the relevant predictors and their coefficients.
* **predictor.splits:** a list of the relevant predictors.

### Step 2: CovCoExpNets pipeline

Once we have a simulated dataset, we can execute the CovCoExpNets pipeline:

```{r}
x = sim.obj$data %>% as.matrix() %>% t()

# We need to normalize the response variable and extract the covariate (eliminate the mean and standard deviation)
y = sim.obj$y %>% normalize() %>% extract2("covariate")

pred.freq = geneFrequency(x, y, seed = 1796)
pred.subset = reduceGenes(pred.freq, mrfa = 0.9)
cvfit = glmnetGenesSubset(x, y, pred.subset, seed = 1796)
pred.relevant = extractModelGenes(cvfit)
```

The predictors selected as relevant by the model can be seen in `pred.relevant`. As we can see, there is no difference from an ordinary execution with biological data.

### Step 3: Evaluation

The most relevant metrics to evaluate are the *precision*, the *sensibility* and the *specificity*. To do so, we use the `calculateSimulationMetrics` function from `CovCoExpNets`:

```{r}
sim.metrics = calculateSimulationMetrics(pred.relevant, sim.obj, "Q1")
print(sim.metrics)
print(paste0("Precisión: ", sim.metrics$Precision %>% round(2)))
print(paste0("Sensitivity: ", sim.metrics$Sensitivity %>% round(2)))
print(paste0("Specificity: ", sim.metrics$Specificity %>% round(2)))
```

## Parameter testing

To test and compare different simulation parameters, we must first generate a dataframe with all possible combinations:

```{r}
N = c(200, 300)
P = c(1000)
Q = c(100, 500)

hp.df = expand.grid(N = N, P = P, Q = Q) %>% filter(Q < P) 
hp.df
```

Then, we can loop through every row and calculate the simulation metrics:

```{r cache = TRUE}
set.seed(1796)
sim.seeds <- foreach(i=1:nrow(hp.df), .combine="c") %do% {sample(1:999999999, 1)}

hp.metrics = foreach(row = 1:nrow(hp.df), .combine = "rbind") %do%{
  hp.row = hp.df[row, ]
  sim.obj = CovCoExpNets::simulateUnivariate(hp.row$N, hp.row$P, hp.row$Q, seed = sim.seeds[i])
  
  x = sim.obj$data %>% as.matrix() %>% t()
  y = sim.obj$y %>% normalize() %>% extract2("covariate")
  
  pred.freq = geneFrequency(x, y, seed = sim.seeds[i])
  pred.subset = reduceGenes(pred.freq, mrfa = 0.9)
  cvfit = glmnetGenesSubset(x, y, pred.subset, seed = sim.seeds[i])
  pred.relevant = extractModelGenes(cvfit)
  
  sim.metrics = calculateSimulationMetrics(pred.relevant, sim.obj, "Q1") %>% 
    dplyr::mutate(Simulation = row, N = hp.row$N, P = hp.row$P, Q = hp.row$Q, .before = "Precision")
  sim.metrics
}

hp.metrics
```

To properly execute this experiment, it is recommended to repeat each simulation a set number of times and calculate the average and standard deviation of the different metrics. It is also possible to calculate the RMSE if the simulated dataset is split in training/testing.

## Bivariate simulations

With the same simulation framework, it is possible to simulate two covarariates at the same time. The main parameters we need are the number of samples `N`, the total number of predictors `P` and how many of them will be relevant to the first response variable `Q1`, to the second response variables `Q2` and the number of common relevant predictors between the two.

```{r}
N = 200
P = 1000
Q1 = 100
Q2 = 75
Qc = 25

sim.obj = CovCoExpNets::simulateBivariate(N, P, Q1, Q2, Qc, distribution = "gaussian", params = list(mean = 1, sd = 1))

names(sim.obj)
```

In the returned object, we will find:

* **data:** the data.frame with the simulated dataset.
* **y1:** the first response variable.
* **y2:** the second response variable.
* **df.coefficients.1:** a data.frame with two rows: the relevant predictors and their coefficients for the first response variable.
* **df.coefficients.1:** a data.frame with two rows: the relevant predictors and their coefficients for the second response variable.
* **predictor.splits:** a list of the relevant predictors for each response variable and the common relevant predictors.
